---
title: "Human Activity"
author: "Ravi S. Karedla"
date: "Monday, November 16, 2015"
output: html_document
---

###Executive Summary
The object of this project is to predict the quality of workout of six individuals using the personal activity data record by http://groupware.les.inf.puc-rio.br/har team. 
To predict the manner in which the exercise was done, a prediction model was built on the test data provided & the model was tested on 20 situations in the testing data, which is also provided. 

###Data Exploration 
The data has recordings for 160 parameter. Many of the parameter have significant number of 'NA' or no data. Henced can be cleaned.

###Data Cleaning
Columns significant number of 'NA' or no data are removed from both the training & testing data

```{r, echo=TRUE}
pmltrain <- read.csv("pml-training.csv")
pmltest <- read.csv("pml-testing.csv")
colrem <- c("X","new_window","num_window",
            "raw_timestamp_part_1","raw_timestamp_part_2","cvtd_timestamp",            
            "kurtosis_roll_belt","kurtosis_picth_belt","kurtosis_yaw_belt",
            "skewness_roll_belt","skewness_roll_belt.1","skewness_yaw_belt",
            "max_roll_belt","max_picth_belt","max_yaw_belt",      
            "min_roll_belt","min_pitch_belt","min_yaw_belt",
            "amplitude_roll_belt","amplitude_pitch_belt","amplitude_yaw_belt",
            "var_total_accel_belt","avg_roll_belt","stddev_roll_belt","var_roll_belt",
            "avg_pitch_belt","stddev_pitch_belt","var_pitch_belt",
            "avg_yaw_belt","stddev_yaw_belt","var_yaw_belt",
            
            "var_accel_arm",
            "avg_roll_arm","stddev_roll_arm","var_roll_arm",
            "avg_pitch_arm","stddev_pitch_arm","var_pitch_arm",
            "avg_yaw_arm","stddev_yaw_arm","var_yaw_arm",
            "kurtosis_roll_arm","kurtosis_picth_arm","kurtosis_yaw_arm",
            "skewness_roll_arm","skewness_pitch_arm","skewness_yaw_arm",        
            "max_roll_arm","max_picth_arm","max_yaw_arm",
            "min_roll_arm","min_pitch_arm","min_yaw_arm",
            "amplitude_roll_arm","amplitude_pitch_arm","amplitude_yaw_arm",
            
            "kurtosis_roll_dumbbell","kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell",
            "skewness_roll_dumbbell","skewness_pitch_dumbbell","skewness_yaw_dumbbell",
            "max_roll_dumbbell","max_picth_dumbbell","max_yaw_dumbbell",
            "min_roll_dumbbell","min_pitch_dumbbell","min_yaw_dumbbell",
            "amplitude_roll_dumbbell","amplitude_pitch_dumbbell","amplitude_yaw_dumbbell",
            "var_accel_dumbbell",
            "avg_roll_dumbbell","stddev_roll_dumbbell","var_roll_dumbbell",
            "avg_pitch_dumbbell","stddev_pitch_dumbbell","var_pitch_dumbbell",
            "avg_yaw_dumbbell","stddev_yaw_dumbbell","var_yaw_dumbbell",
            
            "kurtosis_roll_forearm","kurtosis_picth_forearm","kurtosis_yaw_forearm",
            "skewness_roll_forearm","skewness_pitch_forearm","skewness_yaw_forearm",
            "max_roll_forearm","max_picth_forearm","max_yaw_forearm",
            "min_roll_forearm","min_pitch_forearm","min_yaw_forearm",
            "amplitude_roll_forearm","amplitude_pitch_forearm","amplitude_yaw_forearm",
            "var_accel_forearm",
            "avg_roll_forearm","stddev_roll_forearm","var_roll_forearm",
            "avg_pitch_forearm","stddev_pitch_forearm","var_pitch_forearm",
            "avg_yaw_forearm","stddev_yaw_forearm","var_yaw_forearm")

# update the training set
pmltraincln <- pmltrain[,-which(names(pmltrain) %in% colrem)]

# update the test set
pmltestcln <- pmltest[,-which(names(pmltest) %in% colrem)]
```

###Cross Validation
To build the prediction model, the caret package will be used. The default behaviour of models in caret is re-sampling based testing, this default process is made use of. The prediction model is built on training data and prediction is done on testing data.

###Model Development
Model is built using Random Forest as it is known for its accuracy. (Note the below two lines are comment for this article as run time for RF in my laptop is 4hrs & the intention here is to just show the steps.)
```{r, echo=TRUE}
# library(caret)
# pmlmodel <- train(classe ~ .,method='rf',data=pmltraincln)
```

### Error Estimation
Typically, the out of sample error is calculated using confusionMatrix, but as the 'classe' variable is not provided in the testing sample, error is calculated from model accuracy. The model is viewed by typing 'pmlmodel' in R - 

```{r, echo=TRUE}
# Random Forest 
#
# 19622 samples
#    53 predictor
#     5 classes: 'A', 'B', 'C', 'D', 'E' 
#
# No pre-processing
# Resampling: Bootstrapped (25 reps) 
# Summary of sample sizes: 19622, 19622, 19622, 19622, 19622, 19622, ... 
# Resampling results across tuning parameters:
# 
#   mtry  Accuracy   Kappa      Accuracy SD   Kappa SD   
#    2    0.9927599  0.9908428  0.0009785552  0.001238721
#   29    0.9929300  0.9910583  0.0011823548  0.001497079
#   57    0.9851896  0.9812681  0.0032328961  0.004091092
# 
# Accuracy was used to select the optimal model using  the largest value.
# The final value used for the model was mtry = 29. 
```
As the mtry=29 is the final model, the accuracy=0.99293 i.e., the error=0.007 (which is 'Bond, James Bond' error :) )

###Conclusion
A Random Forest model is built to predict the manner in which the exercise was done. This model was used to predict 20 situations in the testing data.